---
title: "RiboCryptOverview"
date: "`r BiocStyle::doc_date()`"
author: "Michal Swirski & Hakon Tjeldnes"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{RiboCrypt Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8} 
---

# Introduction

Welcome to the `RiboCrypt` package. 
`RiboCrypt` is an R package for interactive visualization in genomics. `RiboCrypt` works with any NGS-based method, but much emphasis is put on Ribo-seq data visualization. 

This vignette will walk you through usage with examples. 

`RibCrypt` currently supports creating interactive browser views for NGS tracks:

1. Coverage plots for RNA-seq or any other type of experiment
2. Sub-codon resolution display of ribo-seq p-shifted footprints. 3 reading frames can be displayed as:
  -lines
  -columns
  -stacked area
3. Coverage profiles smoothing via the use of sliding-window (k-mers)
4. Genome-transcriptome toggling, that is data display in genomic or transcriptomic context.
5. Animating between conditions, replicates, readlengths, sliding window width or any other parameter.

# Introduction to Ribo-seq

If you're not familiar with terms like "p-shifting" or "p-site offset", it's best to walk through ORFikOverview vignette, especially chapter 6 "RiboSeq footprints automatic shift detection and shifting"

https://bioconductor.org/packages/release/bioc/vignettes/ORFik/inst/doc/ORFikOverview.html#riboseq-footprints-automatic-shift-detection-and-shifting




# Loading ORFik experiment

The simplest way to create a window is to use the syntax of programing found in the package ORFik.

```{r eval = TRUE, echo = TRUE, message = FALSE}
library(RiboCrypt) # This package
library(ORFik)     # The backend package for RiboCrypt
```

Now load the experiment and load the coding sequences (cds). Here, we use ORFik experiment data structure, to familiarize yourself with the concept check the ORFikExperiment vignette:

https://bioconductor.org/packages/release/bioc/vignettes/ORFik/inst/doc/ORFikExperiment.html

```{r eval = TRUE, echo = TRUE, message = FALSE}
 df <- RiboCrypt.template.experiment()
 cds <- loadRegion(df, "cds")
 cds # gene annotation
 df # let's look at libraries the experiment contains

```
We can see that the experiment consists of CAGE (5'ends), PAS (3'ends), RNA-seq and Ribo-seq libraries in wild-type and mutant conditions.

ORFik experiment can by subsetted either by index or by column. It will be useful to pick libraries to display.
```{r eval = TRUE, echo = TRUE, message = FALSE}
df[4:6,]
df[which(df$libtype == "CAGE"),]
df[which(df$condition == "Mutant"),]

```
# Creating browser window

Browser window is created with `multiOmicsPlot` functions. The function displays libraries subsetted from experiment (df) from top to down. First, lets have a look at a single library of each type (cds is extended by 30 bases). In the resulting plot you can zoom in and inspect coverage:

```{r eval = TRUE, echo = TRUE, message = FALSE}

  multiOmicsPlot_ORFikExp(extendLeaders(extendTrailers(cds[3], 30), 30), annotation = cds,df = df[c(1,5,9,13),],
                        frames_type = "columns", custom_motif = "CTG")
 
```

From top to down, we see Ribo-seq coverage, displayed as columns with the three frames color-coded according to the bottom panel: the three rectangles represent reading frames with sequence features denoted by vertical lines: white for AUG codons and black for stop codons. Custom motifs can be displayed in purple with custom_motif argument. Middle panel is gene model schematic, note that cds frame is also color-coded to match reading frame.

Now, we can plot just the ribosome footprints (RFPs), with display changed to lines, instead of columns:

```{r eval = TRUE, echo = TRUE, message = FALSE}

  multiOmicsPlot_ORFikExp(extendLeaders(extendTrailers(cds[3], 30), 30), annotation = cds,df = df[which(df$libtype == "RFP")[1],],
                        frames_type = "lines")
 
```

Line display is intuitive, but Ribo-seq coverage tends to be very serrated - lines overlap and the whole picture is blurry. To aid that, you can use kmers argument, applying sliding window (sum or mean) over each frame, decreasing resolution and allowing for clearer separation of the three lines. It's also useful when significantly zoomed out:

```{r eval = TRUE, echo = TRUE, message = FALSE}

  multiOmicsPlot_ORFikExp(extendLeaders(extendTrailers(cds[3], 30), 30), annotation = cds,df = df[which(df$libtype == "RFP")[1],],
                        frames_type = "lines", kmers = 6)
 
```

We can now explore the stacked area reading frame display. It's especially useful for small figure generation: lines or columns may be indistinguishable when plot is reduced to very small dimensions for publication purposes, when stacked area should be well visible even when zoomed-out significantly. Note the camera icon in top-right panel. It allows for static image download. By default it's svg, allowing for vector graphics support and high-quality figure generation.

```{r eval = TRUE, echo = TRUE, message = FALSE}

  multiOmicsPlot_ORFikExp(extendLeaders(extendTrailers(cds[3], 30), 30), annotation = cds,df = df[which(df$libtype == "RFP")[1],],
                        frames_type = "stacks", kmers = 6)
 
```

# Conclusion

RiboCrypt offers interactive NGS profile display with several additional visualization methods designed for Ribo-seq specifically. Other utilities include kmers smoothing and static vector graphics export which for publication-grade figures generation.

